The \project{unRAVE} catalog

### 2016-09-02 article/unrave.tex

\newcommand{\stub}[1]{\textbf{#1}}

@@

\newcommand{\Nstars}{483,330}

@@

\begin{abstract}
The Milky Way may be our greatest laboratory for understanding galaxy formation.
This premise relies on the ability to precisely determine the astrometry and
chemistry for many individual stars, and to use those data to make inferences
about the Galaxy's formation and evolution. On 14 September 2016 the \gaia\ 
mission will release precise astrometry for $\approx2\times10^6$ stars.  
However, a comprehensive catalog of detailed chemical abundances to complement
those data is currently lacking.  The RAdial Velocity Experiment (\rave) survey 
has acquired spectra for 292,036 stars in common with the first \gaia\ data 
release, constituting the largest overlap of any stellar spectroscopic sample.  
Here we perform an independent analysis of \rave\ spectra using an implementation
of \thecannon\ that uses prior probabilities on atomic line formation and stellar
astrophysical parameters.  Our approach yields precise stellar labels at lower 
S/N ratios than most \emph{ab initio} approaches, allowing us to deliver improved
radial velocities, effective temperature $\teff$, surface gravity $\logg$, and 
chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, Ni) for 
\Nstars\ stars.  We validate our results internally from repeat visits, and 
externally against existing high-resolution spectroscopic surveys.  The typical
precision in chemical abundances is X.XX~dex.  We provide spectrophotometric 
distances based on our labels, improving upon the parallax distances available
for the fainter targets in the first \gaia\ data release. \stub{A one-line 
science result based on RAVE/Hipparcos overlap}
\end{abstract}

@@

\rave\ is a magnitude-limited stellar spectroscopic survey of the (nearby) Milky Way,
principally designed to measure radial velocities for up to $10^6$ stars.
Observations were conducted on the 1.2~m UK Schmidt telescope at the Australian 
Astronomical Observatory\footnote{Formerly the Anglo-Australian Observatory} from 
2003--2013.  The large 5.7~degree field-of-view and robotic fibre positioner (6dF)
made for very efficient observing.  Spectra for up to 150 targets were simultaneously
acquired, and when observations concluded in April 2013, a total of 574,630 spectra
had been collected of 483,330 unique objects.

@@

The model we have introduced is generic in the sense that there is no description for how
complex the vectorizing function $\vecv(\ell)$ should be, or how many $K$ labels are used.
Given previous successes using a second-order (quadratic) polynomial model 
\citep{Ness_2015,Ness_2016,Ho_2016,Casey_2016b}, we will first adopt a three-label ($\teff$, $\logg$, [Fe/H]) model for all (giant and main-sequence) stars. Thus, for our first set of
model experiments,
\begin{eqnarray}\label{eq:vectorizer-three-label}
\vecv(\ell_n) & \rightarrow & \left[1, T_{{\rm eff},n}, \logg_n, [{\rm Fe/H}]_n, T_{{\rm eff},n}^2, \logg_n\cdot{}T_{{\rm eff},n}, \feh_n\cdot{}T_{{\rm eff},n}, \logg_n^2, \feh_n\cdot\logg_n, \feh_n^2\right]
\quad ,
\end{eqnarray}

@@

\stub{We trained the model initially with no regularization. Could reproduce general structure of a main-sequence and HRD, but the intricacies of the structure was gone. Model appeared to be too simple. However we can use this model to identify dwarfs/giants,etc. We may revisit this model later}

@@

We constructed a three-label quadratic model for the main-sequence and sub-giant
stars in the \epic\ training set.  In order to set the regularization hyperparameter
$\Lambda$ for this model, we trained 30 models with different regularization strengths,
spaced evenly in logarithmic space between $\Lambda = 10^{-3}$ to $\Lambda = 10^{3}$.
We then performed leave-one-out cross-validation for each model.  Specifically, for 
each star in the training set: we removed the star; trained the model; and then 
inferred labels from the removed star as if it was a test object. We also performed 
leave-one-out cross-validation on an unregularized ($\Lambda = 0$) model, which we 
will use as the basis for comparison.  For every model we calculated the bias and
root-mean-square (RMS) deviation between the input labels and those inferred through
cross-validation.

@@

We used a quadratic vectorizer for the giant star model, and set the regularization
hyperparameter $\Lambda$ in the same way described in Section \ref{sec:ms-model}.
The results are shown in Figure \ref{fig:set-giant-model-hyperparameters}, where
again the enveloped region represents the minimum and maximum change for any label
with respect to the unregularized case.  At the point of maximum improvement 
near $\Lambda = X.XX$, the RMS in all nine labels has decreased by at least $X$\%, 
and up to $X$\%.  Here the regularization strength also produces a sparser matrix
of $\vectheta$, with a $\approx20$\% increase in the fraction of zero-valued 
entries, particularly for the cross-terms (Figure \ref{fig:giant-model-sparsity},
third panel).  Based on decreasing RMS deviation in the labels, we adopt 
$\Lambda = X.XX$ for our giant star model.

@@

\section{Validation experiments}\label{sec:validation}

We have already performed leave-one-out cross-validation on all of our models
as part of the hyperparameter selection.  Here we describe the results from our
internal, external and astrophysical validation checks.

\subsection{Multi-epoch observations}\label{sec:repeat-observations}

The \rave\ survey performed repeat observations for 43,918 stars with time 
intervals ranging from a few hours to up to four years.  This timing was 
constructed to be quasi-logarithmic such that spectroscopic binaries could
be optimally identified. Most of the stars that were observed multiple times
were only observed twice, with thirteen visits being the maximum number 
of observations for any target.  These repeat observations allow us to 
quantify the level of (in)correctness in our formal errors.  For every star
with multiple visits we constructed a high S/N stacked spectrum for that
star, weighted by the inverse variances in the individual visit spectra,
which then served as the basis of comparison to all individual visits.  
Figure \ref{fig:formal-errors-comparison} shows the difference in labels 
between the comparison spectrum and a repeat visit, normalized by their 
formal errors summed in quadrature (e.g., 
$\Delta\logg/\sqrt{\sigma_{\logg,stacked}^2 + \sigma_{\logg,visit}^2}$).
If our measurements were unbiased by S/N and the errors were representative, 
these values should be normally distributed with a zero mean and unit variance.
\stub{Are they? Should we inflate our formal errors?}

% Internal validation:
% --> bootstrap resampling of the training set?

% Identify outliers:
% --> Stars with morphological classification
% --> Some rogue gallery of a few high S/N cases that have clearly gone haywire?

@@

\stub{What abundances are available?}

% This is more relevant for the discussion, not description of the training set.
We note that the while Ti can be measured, its values in both 
APOGEE DR12 and DR13 are known to not follow the expected Galactic trends \citep{Holtzman_2015, Hawkins_2016}.

\stub{Any differences in RAVE abundances?}

% External validation: 
% --> Gaia-ESO Survey
% --> Literature studies presented in Kordopatis+ 2014
% --> Gold standards (incl. exoplanet host star studies): Reddy (2003, 2006), Bensby (2014), Valenti & Fischer (2005)

\subsection{Astrophysical validation}

After verifying that our stellar labels are comparable with other
high-resolution studies, here we verify that the abundance labels we find
are consistent with expectations from astrophysics.  One example is stars in
globular clusters, which show anti-correlations in light element abundances 
\citep[e.g.,][and references therein]{Da_Costa,Carretta_2009}. In the \rave\
survey, \cite{Anguiano_2015} found 70 stars with positions and radial velocities that are
consistent with being members of globular clusters.  We have used that sample
Specifically, this GC sample includes 49 stars belonging to NGC~5139 ($\omega$~Centauri), 11 members of the retrograde
globular cluster NGC~3201, and 10 members of NGC~362.  
\stub{Were any of these stars unintentionally in the training set? \\ KH: Not using my training sets.}
All of these systems are well-studied and exhibit abundance anti-correlations 
in Na--O and Mg--Al \citep{people}.  Other correlations are also present 
(e.g., C--N), but those relationships are less relevant for this work because 
we do not measure C and N abundances here.

@@

@@

% TODO: Replace ZENODO-URL with a data url
\noindent\colorbox{unoffensive-warning}{\parbox{\dimexpr\linewidth-2\fboxsep}{
\noindent{}Source code for this project is available at \texttt{\giturl}\hspace{-0.5em},
and this document was compiled from revision hash \texttt{\githash} in that repository.
Inferred labels and covariance matrices are available electronically with 
associated metadata at \texttt{\url{ZENODO-URL}} \citep{DATA_REPOSITORY}.  
Please note that it is a condition of using these results that the fifth \rave\ 
data release by \citet{Kunder_2016} must (also) be cited, as the work presented 
here would not have been possible without the tireless efforts of the entire 
\rave\ collaboration, past and present.
}}

@@

\begin{figure}[p]
\caption{A H-R diagram showing the training set labels.\label{fig:training-set-hrd}}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figures/hrd-test-set.pdf}
\caption{A `spectroscopic' Hertzsprung-Russell diagram showing the effective temperature $\teff$ and surface gravity $\logg$ for stars in the test set.  Only stars with $\chi_r < 3$ are shown, and each panel shows stars that meet a threshold in $S/N$ ratio.\label{fig:test-set-hrd}}
\end{figure}

@@

\begin{figure*}[p]
\caption{Stellar label comparison between the \project{UNRAVE} catalog and the \project{RAVE} fifth data release \citep{Kunder_2016}.\label{fig:rave-dr5-comparison}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/ges-comparison.pdf}
\caption{Stellar label comparison between the \project{UNRAVE} catalog and the \project{Gaia-ESO Survey} fourth internal data release.\label{fig:ges-dr4-comparison}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/literature-comparison.png}
\caption{Stellar label comparisons between the \project{UNRAVE} catalog and the samples discussed in \citep{Kordopatis_2013}}
\end{figure*}

>>>>

\newcommand{\stub}[1]{{\color{blue} \textbf{#1}}}

@@

\newcommand{\Nspectra}{520,782}
\newcommand{\Nstars}{457,589}

@@

\begin{abstract}
The Milky Way may be our greatest laboratory for understanding galaxy formation.
This premise relies on the ability to precisely determine the astrometry and
chemistry for many individual stars, and to use those data to make inferences
about the Galaxy's formation and evolution. On 14 September 2016 the \gaia\ 
mission will release precise astrometry for $\approx2\times10^6$ stars.  
However, a comprehensive catalog of detailed chemical abundances to complement
those data is currently lacking.  The RAdial Velocity Experiment (\rave) survey 
has acquired spectra for 292,036 stars in common with the first \gaia\ data 
release, constituting the largest overlap of any stellar spectroscopic sample.  
Here we perform an independent analysis of \rave\ spectra using an implementation
of \thecannon\ that uses prior probabilities on atomic line formation and stellar
astrophysical parameters.  Our approach yields precise stellar labels at lower 
S/N ratios than most \emph{ab initio} approaches, allowing us to deliver improved
radial velocities, effective temperature $\teff$, surface gravity $\logg$, and 
chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, Ni) for 
\Nstars\ stars.  We validate our results internally from repeat visits, and 
externally against existing high-resolution spectroscopic surveys.  The typical
precision in chemical abundances is X.XX~dex.  We provide spectrophotometric 
distances based on our labels, improving upon the parallax distances available
for the fainter targets in the first \gaia\ data release. \stub{A one-line 
science result based on RAVE/Hipparcos overlap.}
\end{abstract}

@@

\rave\ is a magnitude-limited stellar spectroscopic survey of the (nearby) Milky Way,
principally designed to measure radial velocities for up to $10^6$ stars.
Observations were conducted on the 1.2~m UK Schmidt telescope at the Australian 
Astronomical Observatory\footnote{Formerly the Anglo-Australian Observatory} from 
2003--2013.  A large 5.7~degree field-of-view and robotic fibre positioner 
made for very efficient observing;  spectra for up to 150 targets could be simultaneously
acquired.  When observations concluded in April 2013, a total of \Nspectra\ useful spectra
had been collected of \Nstars\ unique objects.

@@

The model we have introduced is generic in the sense that there is no description for how
complex the vectorizing function $\vecv(\ell)$ should be, or how many $K$ labels are used.
Given previous successes using a second-order (quadratic) polynomial model 
\citep{Ness_2015,Ness_2016,Ho_2016,Casey_2016b}, we will first adopt a three-label ($\teff$, $\logg$, [Fe/H]) model for all (giant and main-sequence) stars. Thus, for our first set of
model experiments,
\begin{eqnarray}\label{eq:vectorizer-three-label}
\vecv(\ell_n) \rightarrow \left[1, T_{{\rm eff},n}, \logg_n, [{\rm Fe/H}]_n, T_{{\rm eff},n}^2, \logg_n\cdot{}T_{{\rm eff},n}, \feh_n\cdot{}T_{{\rm eff},n}, \logg_n^2, \feh_n\cdot\logg_n, \feh_n^2\right]
\end{eqnarray}

@@

\stub{We trained the model initially with no regularization. Could reproduce general structure of a main-sequence and HRD, but the intricacies of the structure was gone. Model appeared to be too simple.  However we can use this model to identify dwarfs/giants,etc. Rather than go to a very complex model, we opted to use two models: one for the main-sequence and one for the giant branch.}

@@

We constructed a three-label quadratic model for the main-sequence and sub-giant
stars in the \epic\ training set.  In order to set the regularization hyperparameter
$\Lambda$ for this model, we trained 30 models with different regularization strengths,
spaced evenly in logarithmic steps between $\Lambda = 10^{-3}$ to $\Lambda = 10^{3}$.
We then performed leave-one-out cross-validation for each model.  Specifically, for 
each star in the training set: we removed the star; trained the model; and then 
inferred labels from the removed star as if it was a test object. We also performed 
leave-one-out cross-validation on an unregularized ($\Lambda = 0$) model, which we 
will use as the basis for comparison.  For every model we calculated the bias and
root-mean-square (RMS) deviation between the input labels and those inferred through
cross-validation.

@@

We used a quadratic vectorizer for the giant star model, where the quadratic and
cross-terms are expanded in the same way as \ref{eq:vectorizer-three-label}, only
with nine labels.  The regularization hyperparameter $\Lambda$ was set in the same
way described in Section \ref{sec:ms-model}, using the same 30 trials of $\Lambda$.
The results are shown in Figure \ref{fig:set-giant-model-hyperparameters}, where
again the enveloped region represents the minimum and maximum change in RMS label
deviation with respect to the unregularized case.  At the point of maximum improvement 
near $\Lambda = X.XX$, the RMS in all nine labels has decreased by at least $X$\%, 
and up to $X$\%.  Here the regularization strength also produces a sparser matrix
of $\vectheta$, with a $\approx20$\% increase in the fraction of zero-valued 
entries, particularly for the cross-terms (Figure \ref{fig:giant-model-sparsity},
third panel).  Based on the increased model sparsity and decreasing RMS deviation
in the labels, we adopt $\Lambda = X.XX$ for our giant star model.

@@

\section{Validation experiments}
\label{sec:validation}

In addition to the cross-validation tests that we have previously described, 
we conducted a number of internal and external validation experiments to 
test the validity of our results.

\subsection{Multi-epoch observations}
\label{sec:repeat-observations}

The \rave\ survey performed repeat observations for 43,918 stars with time 
intervals ranging from a few hours to up to four years.  This timing was 
constructed to be quasi-logarithmic such that spectroscopic binaries could
be optimally identified. Most of the stars that were observed multiple times
were only observed twice, with thirteen visits being the maximum number 
of observations for any target.  These repeat observations allow us to 
quantify the level of (in)correctness in our formal errors.  For every star
with multiple visits we constructed a high S/N stacked spectrum for that
star, weighted by the inverse variances in the individual visit spectra.

We treated the stacked spectra as normal survey stars.  We inferred labels
using all models and joined the results as per Section \ref{sec:joining-the-models}.  The labels inferred from the stacked spectra served as the basis
of comparison for all individual visits to that object.  Figure 
\ref{fig:formal-errors-comparison} shows the difference in labels 
between the comparison spectrum and a repeat visit, normalized by their 
formal errors summed in quadrature (e.g., 
$\Delta\logg/\sqrt{\sigma_{\logg,stacked}^2 + \sigma_{\logg,visit}^2}$).
If our measurements were unbiased by S/N and the formal errors were 
representative, these values should be normally distributed with a zero 
mean and unit variance.
\stub{Are they? (Probably). How much should we inflate our formal errors, and should this be as a function of S/N? Note that in all Figures and tables,
we show these inflated errors only, as we gauge them to be more representative.}

\stub{Covariances between these parameters are accessible through the matrix of correlation coefficients in the associated results file.}

\subsection{Jackknifing the training set}
\label{sec:jackknife}

\stub{Should we?}

\subsection{Quality control}
\stub{What should we adopt as recommended QC? Show rogue gallery of a few "high s/n cases" where things went haywire?}

@@

\stub{Discuss comparisons with: GES; calibration sources used in \citet{Kordopatis_2013}; and `gold-standard' studies: Reddy, Bensby, Valenti \& Fischer}

\stub{Any differences in RAVE abundances? Corrections or caution required?}

\subsection{Astrophysical validation}
\label{sec:astrophysical-validation}

After verifying that our stellar labels are comparable with other
high-resolution studies, here we verify that the abundance labels we find
are consistent with expectations from astrophysics.  One example is stars in
globular clusters, which show anti-correlations in light element abundances 
\citep[e.g.,][and references therein]{Da_Costa,Carretta_2009}. In the \rave\
survey, \cite{Anguiano_2015} found 70 stars with positions and radial velocities that are
consistent with being members of globular clusters.  We have used that sample
Specifically, this GC sample includes 49 stars belonging to NGC~5139 ($\omega$~Centauri), 11 members of the retrograde
globular cluster NGC~3201, and 10 members of NGC~362.  
All of these systems are well-studied and exhibit abundance anti-correlations 
in Na--O and Mg--Al \citep{people}.  Other correlations are also present 
(e.g., C--N), but those relationships are less relevant for this work because 
we do not measure C and N abundances here.  We note that \emph{none} of these 
cluster stars are present in our training set.

@@

\stub{Chemistry + Kinematics = Stockholm}

\stub{What do the two provide us with?}

\stub{What is one interesting science result should we show based on these abundances and Hipparcos results, which was not known previously?}

@@

% TODO: Replace ZENODO-URL with a data url
\noindent\colorbox{unoffensive-warning}{\parbox{\dimexpr\linewidth-2\fboxsep}{
\noindent{}Source code for this project is available at \texttt{\giturl}\hspace{-0.5em},
and this document was compiled from revision hash \texttt{\githash} in that repository.
Inferred labels, covariance matrices, and relevant metadata are available electronically
at \texttt{\url{ZENODO-URL}} \citep{DATA_REPOSITORY}.  Please note that it is a condition
of using these results that the fifth \rave\ data release by \citet{Kunder_2016} must 
also be cited, as the work presented here would not have been possible without the 
tireless efforts of the entire \rave\ collaboration, past and present.
}}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figures/hrd-train-set.pdf}
\caption{A H-R diagram showing the training set labels.\label{fig:training-set-hrd}}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figures/hrd-test-set.pdf}
\caption{The effective temperature $\teff$ and surface gravity $\logg$ for \rave\ stars after combining labels from the main-sequence and giant star models.  Only stars with $v\sin{i} < 1$~km~s$^{-1}$ and $\chi_{red} < 3$ are shown.  The top three panels show logarithmic density, and bins in the lower three panels are colored by the median metallicity in each bin.\label{fig:test-set-hrd}}
\end{figure}

@@

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/ges-comparison.pdf}
\caption{Stellar label comparison between the \project{unRAVE} catalog and the \ges\ fourth internal data release.\label{fig:ges-stellar-parameters}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/ges-abundances.pdf}
\caption{Individual chemical abundances in \project{unRAVE} compared to \ges.\label{fig:ges-abundances}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/gold-standard-comparison.pdf}
\caption{Stellar parameter comparisons for stars in common between this work and studies using high-resolution, high S/N spectra and \hipparcos\ parallaxes.\label{fig:gold-standard-comparison}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/gold-standard-hrd.pdf}
\caption{HRD of gold standard comparisons.\label{fig:gold-standard-hrd}}
\end{figure*}

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/kordopatis-calibration.pdf}
\caption{Comparison with the calibration (literature) samples used by \citet{Kordopatis_2013} and \citet{Kunder_2016}.\label{fig:kordopatis-calibration}}
\end{figure*}

\begin{figure}[p]
\includegraphics[width=0.5\textwidth]{figures/gce.pdf}
\caption{Giant star abundances ([X/Fe]) with respect to [Fe/H], showing the Galactic chemical evolution trend for individual elements.\label{fig:gce}}
\end{figure}
